---
title: "DataPreprocessing"
author: "Geo"
date: "16 December 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

install.packages("ggplot2movies")  # one-time only - do this if you have not done so already
library(ggplot2movies)  # once per R session
data(movies)

install.packages("plyr")  # one-time only
library(plyr)  # once per R session
data(baseball)

data(ozone)

install.packages("ggplot2")  # one-time only
library(ggplot2)  # once per R session
data(diamonds)

  
install.packages("robustHD")
library("robustHD")

library(MASS)

library(reshape2)

data(tips)
```
is.na

complete.cases

na.omit

na.rm --> argument

```{r}
mean(movies$length)

mean(movies$budget)


mean(movies$budget,na.rm = TRUE)

mean(is.na(movies$budget))

moviesNONA = na.omit(movies)
qplot(rating,budget,data = moviesNONA, size = I(1.2)) + stat_smooth(color = 'red', size = I(2), se = F)

```


Dealing with outlers

Truncating
Remove outlier values
Winsorization
shrink outliers to border of main part of data
Robustness
Choose robust procedure for analysis

```{r}
library(robustHD)
#Create the data to be Winsorized:
originalData <- c(1000.0000000  ,  1.8410249 ,  -0.7923505 ,   0.1776514  ,  0.4002135)
originalData <- c(1000,rnorm(10))
#Print the data to be Winsorized:
print(originalData[1:5])

#TODO winsorize the data:
print(winsorize(originalData))

?winsorize
```


```{r}
library(dplyr)
install.packages('dplyr')
```
##dplyr functions

* filter
– select rows

* select
– select columns

* arrange
– reorder rows

* mutate
– create new columns

* rename
– rename columns
 
group_by
and 
summarize
- aggregate


## Filter
```{r}
df1 <-filter(mtcars,cyl==4, mpg > 20)
df2 <-filter(mtcars, cyl==4 & mpg > 20) 
df3 <-filter(mtcars,cyl %in% c(4, 6) | am ==1)
```

## Select

df1 <-select(mtcars, mpg,cyl, wt)
df2 <-select(mtcars, mpg:qsec, carb)
df3 <-select(mtcars, -am,-carb)



## arrange

df1 <- arrange(mtcars, desc(cyl))
df2 <- arrange(mtcars, cyl, mpg)
df3 <- arrange(mtcars, cyl, desc(mpg))


## mutate

df1 <-mutate(mtcars,power = disp * hp,am = factor(am,levels=c(0, 1),
labels = c("automatic", "manual")))

## rename

df  <- rename(mtcars,displacement = disp,transmission = am)

## group by summarize


df <- group_by(mtcars, cyl, gear)
  
df2 <- summarise(df,disp_n = mean(disp), disp_sd = sd(disp))


df <- group_by(mtcars, cyl, gear)

df2 <- summarise_all(df,funs(mean))

df2 <- summarise_all(df,funs(min,max))

## putting it all together

df <-  select(mtcars, cyl, disp, mpg) %>%
filter(mpg > 20) %>% arrange(cyl, desc(mpg))

## Calculating percentages


mtcars %>%  group_by(cyl) %>% summarise(n = n()) %>% mutate(pct= n/sum(n))



as.data.frame(mtcars%>%  group_by(cyl) %>%summarise(n = n()) %>%mutate(pct =paste0(round(100 * n/sum(n), 0), "%")))


as.data.frame(mtcars%>%  group_by(cyl, gear) %>%summarise(n = n()) %>%mutate(pct =paste0(round(100 * n/sum(n), 0), "%")))


## Windows functions (min_rank)


# what are the 2 automatic transmission cars and 
# 2 manual transmission cars that have the lowest gas mileage?

mtcars$name <- row.names(mtcars)
mtcars%>%  group_by(am) %>% filter(min_rank(mpg) <= 2) %>% 
select(name, am, mpg) 

# have the highest gas mileage?

mtcars %>%  group_by(am) %>% filter(min_rank(desc(mpg)) <= 2) %>% 
select(name, am, mpg)






