---
title: "GapminderWrangling"
author: "Geo"
date: "16 December 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#install.packages('gapminder') # instead of reading in the csv

install.packages('gapminder')
library(gapminder) # this is the package name
library(dplyr)
  
str(gapminder)
```{r}
filter(gapminder, lifeExp < 29 & country %in% c("Mexico", "Afghanistan") )
filter(gapminder, country == "Mexico" )
filter(gapminder, country %in% c("Mexico", "Afghanistan"))


?filter
gapminder[gapminder$lifeExp < 29, ] 
subset(gapminder, country == "Mexico")
```
## pipe operator

```{r}
gapminder %>% head()
```

This is equivalent to head(gapminder). This pipe operator takes the thing on the left-hand-side and pipes it into the function call on the right-hand-side – literally, drops it in as the first argument

```{r}
gapminder %>% head(3)


```


# Use dplyr::select() to subset the data on variables or columns

```{r}
select(gapminder, year, lifeExp)

gapminder %>% 
  select(year, lifeExp) %>% 
  head(4)
```



```{r}
gapminder %>%
  filter(country == "Cambodia") %>%
  # select(country, year, pop, gdpPercap) ## entering 4 of the 6 columns is tedious
  select(-continent, -lifeExp) # you can use - to deselect columns
```
In typical base R,
```{r}
gapminder[gapminder$country == "Cambodia", c("country", "year", "pop", "gdpPercap")]
```
# A nicer way
```{r}
subset(gapminder, country == "Cambodia", select = c(country, year, pop, gdpPercap))
```
## Use mutate() to add new variables

```{r}
gapminder %>% 
  mutate(gdp = pop * gdpPercap)
```
## Exercise: add this new gdp column to our Cambodia example above.

```{r}
# Solution
gapminder %>%
  filter(country == "Cambodia") %>%
  select(-continent, -lifeExp) %>%
  mutate(gdp = pop * gdpPercap) %>%
  mutate(year = year)
```
## group_by and summarize
# calculate the mean gdp across all years

```{r}
gapminder %>%
  filter(country == "Cambodia") %>%
  select(-continent, -lifeExp) %>%
  mutate(gdp = pop * gdpPercap) %>%
  group_by(country) %>%
  summarize(mean_gdp = mean(gdp)) %>%
  ungroup() # if you use group_by, also use ungroup() to save heartache later

```
## For all countries 

```{r}
gapminder %>%
  select(-continent, -lifeExp) %>%
  mutate(gdp = pop * gdpPercap) %>%
  group_by(country) %>%
  summarize(mean_gdp = mean(gdp)) %>%
  ungroup() # if you use group_by, also use ungroup() to save heartache later
```


## arrange

```{r}
gapminder %>%
  filter(year == '1957') %>%
  arrange(desc(gdpPercap))
```

```{r}
gapminder %>%
  filter(year == '1957') %>%
  arrange(desc(gdpPercap)) %>%
  head(1) %>%
  select('country')
```

### tidyr overview

Data must be reshaped for it to become tidy data
   - turn columns into rows (gather()),
   - turn rows into columns (spread()),
   - turn a character column into multiple columns (separate()),
   - turn multiple character columns into a single column (unite())
   
```{r}
#install.packages("tidyr")
library("tidyr") # warning messages are OK
```


```{r}
gap_wide <- read.csv('gapminder_wide.csv')

head(gap_wide)

str(gap_wide)
```

?gather

## converting columns to key-value pairs using gather

```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values)
```
```{r}
str(gap_long)
head(gap_long)
tail(gap_long)
```
```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         3:38)  # could also do -1, -2: 'not column one, not column two
str(gap_long)
head(gap_long)
tail(gap_long)
```
## Better way: identify the columns by name.

```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         dplyr::starts_with('pop'),
         dplyr::starts_with('lifeExp'),
         dplyr::starts_with('gdpPercap'))
str(gap_long)
head(gap_long)
tail(gap_long)
```

## this is another way
```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         -continent, -country)
str(gap_long)
head(gap_long)
tail(gap_long)
```

?separate

```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         -continent, -country) %>%
  separate(obstype_year,
           into = c('obs_type','year'),
           sep="_")
```
```{r}
str(gap_long)
head(gap_long)
tail(gap_long)
```


## Exercise: Using gap_long, calculate the mean life expectancy, population, and gdpPercap for each continent. Hint: use the group_by() and summarize() functions we learned in the dplyr lesson


# solution 
gap_long %>% 
  group_by(continent, obs_type) %>%
    summarize(means = mean(obs_values))
    
    
## From long to intermediate format with spread()
?spread

```{r}
gap_normal <- gap_long %>% 
  spread(obs_type, obs_values)
```



```{r}
dim(gap_normal)
dim(gapminder)
names(gap_normal)
names(gapminder)
```
## Exercise: reorder the columns in gap_normal to match gapminder.

```{r}
# Solution

# one way with dplyr 
gap_normal <- gap_normal %>%
  select(country, continent, year, lifeExp, pop, gdpPercap)

# another way with base R
gap_normal <- gap_normal[,names(gapminder)]
```
```{r}
  all.equal(gap_normal,gapminder)
```

```{r}
head(gap_normal)
head(gapminder)
```

```{r}
gap_normal <- gap_normal %>% arrange(country,continent,year)
all.equal(gap_normal,gapminder)
```
         
```{r}
str(gap_normal)
str(gapminder)
```
   
## Exercise: convert gap_long all the way back to gap_wide
?unite
      
```{r}
head(gap_long) # remember the columns

gap_wide_new <- gap_long %>% 
  # first unite obs_type and year into a new column called var_names. Separate by _
  unite(col = var_names, obs_type, year, sep = "_") %>% 
  # then spread var_names out by key-value pair.
  spread(key = var_names, value = obs_values)
str(gap_wide_new)
```
```{r}
aggregate(lifeExp ~ continent, gapminder, median)
```

```{r}
gapminder %>%
 filter(year == 2007, country %in% c("Kenya", "Peru", "Syria")) %>%
 select(country, continent) %>% 
 left_join(country_codes)
```

     
  
Multi-table joins

    Learning objectives

        Focus on the third tidy data principle
            Each variable forms a column.
            Each observation forms a row.
            Each type of observational unit forms a table.
        Be able to use dplyr’s join functions to merge table
        
        
```{r}
set.seed

x <- data.frame(key= LETTERS[c(1:3, 5)], 
                value1 = sample(1:10, 4), 
                stringsAsFactors = FALSE)

```

```{r}
y <- data.frame(key= LETTERS[c(1:4)], 
                value2 = sample(1:10, 4), 
                stringsAsFactors = FALSE)
```
```{r}
inner_join(x, y, by = "key")
```

   
         